Accelerating Traffic Simulation and Its Real-Time Challenges

### 2012-09-12 draft.tex
completed the first draft

\author {
Manato Hirabayashi, Shinpei Kato, Masato Edahiro\\
\textit{Department of Information Engineering}\\
\textit{Nagoya University}\\
\and
Yuki Sugiyama\\
\textit{Department of Physics}\\
\textit{Nagoya University}\\
}

@@

\begin{abstract}
PLACEHOLDER.
\end{abstract}

@@

We assume the Compute Unified Device Architecture (CUDA) for GPU
programming.
In CUDA, a unit of pieces of code that is launched on the GPU is called
a \textit{kernel}.
The kernel is typically composed of multiple parallelized
\textit{threads} that accelerate computations.
A unit of threads is called a \textit{block} where most computing
resources are shared.
A collection of blocks for the corresponding kernel is called a
\textit{grid}. 
The maximum number of threads that an individual block can contain is
defined by the GPU architecture.

GPU applications use a set of the API supported by CUDA.
We typically take the following steps to use the GPU: (i) allocate space
to device memory, (ii) copy data to the allocated device memory space,
(iii) launch the program on the GPU, (iv) copy resultant data back to
host memory, and (v) free the allocated device memory space.

@@

PLACEHOLDER.

@@

Note that Step 3 exploits a lot of loop procedures to derive the
locations and velocities of agents.
This part can be accelerated by the GPU.

@@

Figure~\ref{fig:sample-in-cpu} shows a brief overview of the
Sample-in-CPU scheme.
This scheme also has several alternatives depending on how many kernels
need to be launched on the GPU in each period. 
Suppose that we want to offload $n$ pieces of \textit{for} loops on to
the GPU.
We may return to the CPU $n$ times in totall, that is, return every time
one \textit{for} loop breaks, or otherwise we may just return once when
all the $n$ \textit{for} loops end.
This is a design decision, and is also dependent on the GPU architecture
and the parallelization structure.
In general, parallelized threads need to be synchronized when moving
across basic blocks.
Specifically, when moving to the next \textit{for} loop executed in
parallel, all the threads relevant to this parallel computing procedure
must synchronize with each other.
However, the maximum number of threads that can be synchronized on the
GPU is often limited.
As of 2012, for example, NVIDIA's GPU architectures limit the number of
such threads to 1024 or less~\cite{NVIDIA_Kepler}.
Our implementation therefore forces the program to return to the CPU
every time one \textit{for} loop breaks.
Note that this is not a conceptual limitation of GPU computing, but is a
current limitation of hardware.
We believe that this limitation would be removed or mitigated in
next-generation GPU architectures.

@@

Figure~\ref{fig:eval_benefit} demonstrates the performance benefit of
the GPU under the current hardware limitation.
The Sample-in-CPU scheme herein exploits synchronization between the CPU
and the GPU after every \textit{for} loop to overcome the limitation that
CUDA threads cannot synchronize beyond blocks.
Despite of this restricted implementation, our GPU implementation
outperforms the original CPU implementation by about five times in
simulation time, when the number of agents exceeds a scale of 1000.
This evinces an affinity of the GPU and parallel computing for traffic
simulation.
It is also interesting to see that the CPU implementation has a better
performance for a small number of agents.
Thus, the overhead of communication between the CPU and the GPU has
non-trivial impact if the achievable parallelism is not sufficient.
Another notable observation is that the diffence in performance of the
CPU and the GPU implementations is saturated when the number of agents
reaches 1000.
This saturation implies the hardware potential of the NVIDIA GeForce GTX
560 Ti graphics card used in this evaluation.
Using high-end graphics cards, the difference in performance may not be
saturated at the same scale.

@@

It is very notable that the Sample-in-GPU scheme improves the simulation
time by a factor of 1000 as compared to the original CPU implementation.
This result encourage suport for thread synchronization among blocks in
CUDA programming.
There are also several interesting observations obtained in this
experiment.
Comparing the Sample-in-CPU and the Sample-in-CPU-Sync schemes, it turns
out that the cost of downloading intermediate results from the GPU to
the CPU is becoming trivial as the scale of simulation increases.
Hence, a common argument of ``host-device data communication in GPU
programming is expensive'' does not apply to large-scale traffic
simulation.
Another issue of concern often discussed in GPU programming is that
a single-thread performance of GPUs is weak as compared to that of CPUs.
This is true in general, but our experimental result shows that the
performance loss caused by pushing most pieces of code into the GPU in
the Sample-in-GPU scheme is not signficant even for a small number of
agents.
This observation leads to some conclusion that the Sample-in-GPU scheme
would be the best choice for any scale of traffic simulation.

@@

PLACEHOLDER.

Imprecise Computation~\cite{Lin1987}.
Gdev~\cite{Kato2012}.

@@

PLACEHOLDER.

>>>>

\author {
Manato Hirabayashi, Shinpei Kato and Masato Edahiro\\
\textit{Department of Information Engineering}\\
\textit{Nagoya University}\\
\and
Yuki Sugiyama\\
\textit{Department of Physics}\\
\textit{Nagoya University}\\
}

@@

\begin{abstract}
 Traffic simulation is a growing domain of computational physics.
 There are many applications in our industry and life that benefit from
 traffic simulation.
 A core challenge of this research domain, however, is posed due to
 an unbounded scale of computation.
 In this paper, we explore an advantage of using the graphics processing
 unit (GPU) to support its computation.
 Two schemes are presented to study how to maximize GPU
 performance in this context.
 Our evaluation using a commodity graphics card shows that our GPU
 implementation improves simulation time by about five times over the
 original CPU implementation.
 We also demonstrate by experiments that additional orders-of-magnitude
 improvements could be achieved by overcoming the current hardware
 limitation.
\end{abstract}

@@

We assume the Compute Unified Device Architecture (CUDA) for GPU
programming.
In CUDA, a unit of pieces of code that is launched on the GPU is called
a \textit{kernel}.
The kernel is typically composed of multiple \textit{threads} that
executes the code in parallel.
A unit of threads that are co-scheduled by hardware is called a
\textit{block}, while a collection of blocks for the corresponding
kernel is called a \textit{grid}.  
The maximum number of threads that can be contained by an individual
block is defined by the GPU architecture.

CUDA programs use a set of the application programming interface (API)
functions to control the GPU.
We usually take the following steps to use the GPU: (i) allocate space
to device memory, (ii) copy data to the allocated device memory space,
(iii) launch the program on the GPU, (iv) copy resultant data back to
host memory, and (v) free the allocated device memory space.

Our computing platform is composed of a single CPU and GPU.
The traffic simulation program is a single process from the CPU point of
view.
Hence, kernels are launched from the CPU sequentially, while each kernel
is executed by parallelized threads on the GPU.

@@

The motivation to study traffic flow is to understand the property of
traffic flow, mainly why traffic congestion occurs.
The study of traffic flow has a long history, and many engineering and
physical models exist.
This paper is particularly focused on the OV model~\cite{Bando1995}.
In this model, a driver is supposed to maintain an optimal velocity that
depends on the driver's headway.
Its motion is given by Equation~\eqref{eqn:ov}.
The optimal velocity function introduced in this equation is represented
by the relationship between the driver's headway and velocity.

The OV model has a simple form with a single parameter of sensitivity
$a$.
This simplicity allows us to understand the mechanisms of phenomena, and
also predict the results of experiments.
The disadvantage of this model, on the other hand, is a little allowance
for control systems to turn up the parameter.
There is indeed a need for collaborations between the research domains
of engineering and physics.
To begin with, however, we restrict our attention to the OV model.
More details are found in~\cite{Bando1995}.

The traffic simulation program based on the OV model is straightforward,
as the model itself has a simple form:

@@

The computation in Step 3 exploits a lot of loop procedures to derive
locations and velocities of agents.
This part is the most compute-intensive, and thus can be accelerated by
the GPU.

@@

Figure~\ref{fig:sample-in-cpu} shows a brief overview of the
Sample-in-CPU scheme.
This scheme also has several alternatives depending on how many kernels
need to be launched on the GPU in each period. 
Suppose that we want to offload $n$ pieces of \textit{for} loops on to
the GPU.
We may return to the CPU $n$ times in total, that is, return every time
one \textit{for} loop breaks, or otherwise we may just return once when
all the $n$ \textit{for} loops end.
This is a design decision, and is also dependent on the GPU architecture
and the parallelization structure.
In general, parallelized threads need to be synchronized when moving
across basic blocks.
Specifically, when moving to the next \textit{for} loop executed in
parallel, all the threads relevant to this parallel computing procedure
must synchronize with each other.
However, the maximum number of threads that can be synchronized on the
GPU is often limited.
As of 2012, for example, NVIDIA's GPU architectures limit the number of
such threads to 1024 or less~\cite{NVIDIA_Kepler}.
Our implementation therefore forces the program to return to the CPU
every time one \textit{for} loop breaks.
Note that this is not a conceptual limitation of GPU computing, but is a
current limitation of hardware.
We believe that this limitation would be removed or mitigated in
next-generation GPU architectures.

@@

Figure~\ref{fig:eval_benefit} demonstrates the performance benefit of
the GPU under the current hardware limitation.
The Sample-in-CPU scheme herein exploits synchronization between the CPU
and the GPU after every \textit{for} loop to overcome the limitation that
CUDA threads cannot synchronize beyond blocks.
Despite of this restricted implementation, our GPU implementation
outperforms the original CPU implementation by about five times in
simulation time, when the number of agents exceeds a scale of 1000.
This evinces an affinity of the GPU and parallel computing for traffic
simulation.
It is also interesting to see that the CPU implementation has a better
performance for a small number of agents.
Thus, the overhead of communication between the CPU and the GPU has
non-trivial impact if the achievable parallelism is not sufficient.
Another notable observation is that the difference in performance of the
CPU and the GPU implementations is saturated when the number of agents
reaches 1000.
This saturation implies the hardware potential of the NVIDIA GeForce GTX
560 Ti graphics card used in this evaluation.
Using high-end graphics cards, the difference in performance may not be
saturated at the same scale.

@@

It is very notable that the Sample-in-GPU scheme improves the simulation
time by a factor of 1000 as compared to the original CPU implementation.
This result encourage support for thread synchronization among blocks in
CUDA programming.
There are also several interesting observations obtained in this
experiment.
Comparing the Sample-in-CPU and the Sample-in-CPU-Sync schemes, it turns
out that the cost of downloading intermediate results from the GPU to
the CPU is becoming trivial as the scale of simulation increases.
Hence, a common argument of ``host-device data communication in GPU
programming is expensive'' does not apply to large-scale traffic
simulation.
Another issue of concern often discussed in GPU programming is that
a single-thread performance of GPUs is weak as compared to that of CPUs.
This is true in general, but our experimental result shows that the
performance loss caused by pushing most pieces of code into the GPU in
the Sample-in-GPU scheme is not significant even for a small number of
agents.
This observation leads to some conclusion that the Sample-in-GPU scheme
would be the best choice for any scale of traffic simulation.

@@

According to the experimental results, it is highly desired to use the
Sample-in-GPU scheme, if available.
It provides an orders-of-magnitude improvement in simulation time.
The downside of this scheme at the moment is a lack of functionality in
thread synchronization due to the current hardware limitation.
Simulation results are never precise unless thread synchronization
operates correctly.
Although we have demonstrated that the error in simulation results
caused by a lack of thread synchronization is acceptable in our
experimental setup, the range of error is never bounded as depicted in
Figure~\ref{fig:eval_error}.
Hence, the availability of this scheme depends highly on practical
setups.

We consider that the problem of this scheme may be relevant to the
Imprecise Computation model~\cite{Lin1987}.
The Sample-in-CPU scheme is precise-but-slow, while the Sample-in-GPU
scheme is imprecise-but-fast.
If we are allowed to employ multiple GPUs in the system, we may run the
two types of simulation programs on different GPUs concurrently.
At some point of time $t$, the simulation result produced by the
Sample-in-GPU scheme represents a scenario way ahead of that produced by
the Sample-in-CPU scheme; the question is how precise it is.
Figure~\ref{fig:eval_error} demonstrates that the error in the
simulation result produced by the Sample-in-GPU scheme is getting larger
as time goes by.
If we can find such a time duration that causes the value of error to
fall out the acceptable range for practical use, we define this time
duration to be $\Delta t$.
Suppose that the two simulation programs based on the Sample-in-GPU and
the Sample-in-CPU schemes, respectively, are running concurrently.
At every interval of time $\Delta t$, we reset the simulation result
produced by the Sample-in-GPU scheme, and input the latest precise
simulation result produced by the Sample-in-CPU scheme during the same
interval.
By this means, the value of error in the simulation result produced by
the Sample-in-GPU scheme never falls out the acceptable range, while we
still benefit from its simulation speed within an internal of time
$\Delta t$.
The value of $\Delta t$ can be determined adaptively based on the
Imprecise Computation model.

So far we have studied a single process of traffic flow.
In reality, however, there are multiple processes of traffic flow.
Real-time feedback from the real road conditions may also be
additionally required to deploy traffic simulation.
All these pieces of simulation must meet the quality-of-service (QoS)
requirement for users.
Now, this turns into a real-time systems problem.
We have developed a lot of solutions to use the GPU in real-time
systems~\cite{Kato2011_3, Kato2011_2, Kato2011_1, Kato2012}.
We believe that some of these work could contribute to GPU-accelerated
real-time traffic simulation.

@@

In this paper, we have presented GPU-accelerated traffic simulation
based on the OV model.
The current hardware limitation of the GPU architecture caps
improvements in performance to a factor of ten or less, as compared to
the original CPU implementation.
However, we demonstrated that additional orders-of-magnitude
improvements could be achieved by removing that limitation.
We also discussed a conceivable solution to cope with that limitation,
using a concept of the Imprecise Computation model.

In future work, we extend our scheme to multidimensional OV models.
This extention allows our research to be applied to a broader range of
problems beyond freeways traffic simulation.
We also plan to consider a specific approach to the integration of the
Imprecise Computation model into our scheme in order to facilitate
deployment of traffic simulation in the real world.
Another interesting direction in this line of work is to coordinate with
engineering-based simulation such as AutoMatrix~\cite{Mangharam2011}.
